Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity \end{large}

### 2017-03-06 draft.tex
.

\textbf{Short Term Plasticity:} A short term plasticity (STP) mechanism acting on recurrent excitatory connections was implemented as presented in \cite{Markram_STP} as an additional stabilization of network activity. It modulates the effective synaptic weights by multiplying the value stored in the weight matrix $w_{ji}$ by two dynamic variables $x$ and $u$, $w^{\rm effective}_{ji} = w_{ji}\cdot x \cdot u$, each synapse owning a pair $(x,u)$. The dynamics of these variables are given by:

\begin{equation}
\dot{x} = \frac{1-x}{\tau_d},\; \dot{u} = \frac{U-u}{\tau_f}
\label{STP_dynamics1}
\end{equation}
Each presynaptic spike furthermore causes a change of $x$ and $u$ by
\begin{equation}
x \rightarrow x - x\cdot u,\; u \rightarrow u + U(1-u)
\label{STP_dynamics2}
\end{equation}
If no spikes arrive the system rests at $x\cdot u = U$. Otherwise, depending on the choice of $\tau_d$ and $\tau_f$, one can achieve a weight modulation that is dominated by potentiation ($\tau_f \gg \tau_d$) or depression ($\tau_f \ll \tau_d$). We chose $U=0.04$, $\tau_d = 0.5s$ and $\tau_f = 2s$ as a rough approximation of the values that were experimentally observed \cite{Markram_STP}, giving it a tendency towards potentiation. However, one should keep in mind that for $U\in [0,1]$, $x\cdot u \in [0,1]$ always holds, thus the factor $x\cdot u$ has a generally diminishing effect. For our choice of variables for example, a Poisson input with a constant rate achieves the best synaptic transmission at a rate of $\sim 4.5 Hz$, corresponding to $x\cdot u \approx 0.2$. \textit{Potentiation} in this context refers to the fact that stronger input strengthens synaptic transmission \emph{compared} to close to zero incoming spikes.

@@

\newpage
\section{Results} \label{results}
We first present the results of a comparison between features of the network activity in both variants in section \ref{activ_analys}, since our main goal of the implementation of diffusive homeostasis was to allow the network to develop a broader distribution of firing rates across excitatory neurons compared to the original version of homeostasis. This is followed by an analytic discussion of an instability we observed within the diffusive homeostatic feedback loop, see Section \ref{theor_osc}. Furthermore, we compare topological features of the network under the influence of diffusive and non-diffusive homeostasis in section \ref{Network_Topology_Section}.

\subsection{Activity Analysis} \label{activ_analys}
We set the time constant of threshold adaptation in the diffusive homeostasis to $\mathrm{2500\, ms}$ on a first attempt, as given by \cite{Sweeney_Paper}. Switching between non-diffusive/diffusive homeostasis after 750 seconds, Figure \ref{full_sim_osci} shows the resulting dynamics of the population activity of the excitatory and inhibitory group (panel A/A*), the average $\rm NO$ concentration at the excitatory neurons' positions (panel B/B*) and the average firing threshold (panel C/C*) within the excitatory group (inhibitory threshold was fixed, see table \ref{LIF_neuron_params}). Both homeostatic mechanisms managed to keep the excitatory population activity in the desired range of 3 Hz. However, what might appear to be slightly faster and stronger random fluctuations in panel A-C were in fact regular oscillations across all three depicted variables, as can be seen in panel A*-C*, showing the dynamics in a smaller time window. While the oscillation amplitudes undergo a rather unpredictable course, the frequency remains at a constant level of $\simeq 0.5\, Hz$. This feature is also illustrated by the fact that the power spectra depicted in Figure \ref{Power_Spec_without_Analysis} have a peak at the frequency that appeared to be dominant in Figure \ref{full_sim_osci}. The fact that the mean over all power spectra of excitatory thresholds differs from the power spectrum of the mean of these thresholds in its amplitude indicates that not all thresholds oscillate at the same phase. The overall shapes of both spectra are still similar, though.
\begin{figure}
\begin{center}
%% oscillating_activity_plot.py
\includegraphics[width=\textwidth]{../../plots/diff_hom/rate_NO_th_compl.png}
\end{center}
%\includegraphics[width=\textwidth]{../../plots/diff_hom/rate_NO_th_closeup.png}
\caption{A/A*: Mean of excitatory (blue) and inhibitory (red) neuronal firing rate. B/B*: NO concentration averaged over readouts at excitatory neurons' positions. C/C*: Average excitatory firing threshold. Non-diffusive homeostasis was used for $\mathrm{0-700\, s}$ , diffusive h. for $\mathrm{700-1500\, s}$. Target activity was $\mathrm{3\, Hz}$.}
\label{full_sim_osci}
\end{figure}
\begin{figure}
\begin{center}
%%% plot_pow_spec.py
\includegraphics[width=0.7\textwidth]{../../plots/power_spec/power_spectrum_sim_only.png}
\caption{Blue: Mean of power spectrum over all excitatory thresholds. Red: Power spectrum of mean over all excitatory thresholds.}
\label{Power_Spec_without_Analysis}
\end{center}
\end{figure}

@@

Dynamical models of interconnected excitatory and inhibitory groups can exhibit sustained oscillations of activity \cite[p.~270]{Theor_Neur_Dayan} on a larger, rate-based level. As a simplest model, one can describe the dynamics of the excitatory and inhibitory populations as a two-dimensional dynamical system, representing the respective population rates $r_e$ and $r_i$

@@

$\mathrm{\langle w_{AB} \rangle,\, \langle N_{\mathrm{in},AB} \rangle}$ and $\mathrm{\langle r_A \rangle}$ denote the mean weight of a synapse connection population $\mathrm{A}$ with population $\mathrm{B}$, the mean number of incoming connections at a neuron in population $\mathrm{B}$ and the mean firing rate in population $\mathrm{A}$ (which makes it necessary to assume changes within the firing rate to be slow enough for a quasi-stationary description of the activity).

@@

One thus needs to incorporate the factor $\mathrm{x_{st.}\cdot u_{st.}}$ by means of the previously described mean input and the contribution to the total variance of the membrane noise when trying to find the fixed point of equations \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2}.

All these aspects taken together make it impossible to find an analytic expression for the fixed point of Equation \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2}. Running a simulation of \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2}, including all aforementioned approximations results in a stable configuration. We set the relevant parameters to their predefined values for this simulation and additionally extracted the mean threshold of the excitatory population which was set by the non-diffusive homeostasis, see C/C* in Figure \ref{full_sim_osci}, yielding a threshold of $\mathrm{-56.963\, mV}$. The results are depicted in Figure \ref{Dyn_Rate_Approx}. The system settled to a fixed point of $\mathrm{(r_{0,e},r_{0,i}) = (2.992\, Hz, 7.144\, Hz)}$, which is quite close to $\mathrm{(3\, Hz, 6.768\, Hz)}$, which is the mean excitatory and inhibitory firing rate between $\mathrm{100-750\, s}$ of the full simulation (i.e., during non-diffusive homeostasis).
\begin{figure}
%% fixed_point_net.py
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/approx_rates_EI.png}
\end{center}
\caption{Dynamics of equations \ref{2d-exc_inh_rate_1} (blue line) and \ref{2d-exc_inh_rate_2} (red line). Dotted lines mark the mean frequencies that were present in the full spiking network. Excitatory: 3 Hz, Inhibitory: 6,768 Hz. \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2} converged to 2.992 Hz and 7.144 Hz respectively.}
\label{Dyn_Rate_Approx}
\end{figure}
In principle, one could now further quantify the stability of the found fixed point by calculating the Jacobian matrix at $\mathrm{(r_{0,e},r_{0,i})}$. For the sake of the initial question, though, namely the source of oscillation, it is sufficient at this point to restrict oneself to a preliminary result: The recurrent network as such (without homeostasis) is stable under the given choice of parameters, assuming validity of the described rate model. Furthermore, on an empirical basis, the occurrence of oscillations appears to be rather dependent on the choice of homeostasis since all parameters of the simulation were kept unchanged upon the transition after 750 s, apart from the homeostatic mechanism itself. Therefore, we present in the following section our further analysis of the dynamics underlying the diffusive homeostatic feedback.

@@

Figure \ref{nNOS_approx_plot} compares the approximation given by \eqref{nNOS_single_sol_approx2} to the full NO production model (equations \eqref{Ca_dyn} and \eqref{nNOS_dyn}). Spikes were drawn from a Poisson process at a rate of 3 Hz. The simplified model fits very well for sufficiently isolated spike events, as predicted. One can observe a slightly worse but acceptable agreement for the rare event of two subsequent spikes appearing very close to each other, as seen in Figure \ref{nNOS_approx_plot} at approximately $\rm 1.5\, s$.

@@

\caption{Time course of nNOS(t) with Poisson spiking at 3 Hz. The full simulation (blue, see equations \eqref{Ca_dyn},\eqref{nNOS_dyn}) is well fitted by the simplified model (red, see equation \eqref{nNOS_single_sol_approx2}). Top panel is a closeup of the first spike event. Dashed lines mark the time span shown in the closeup.}

@@

Returning to the question of how to describe NO synthesis in a simplified form, we note that according to equation \eqref{nNOS_single_sol_approx2}, a single spike causes the release of NO by an amount of $\frac{{Ca^{2+}_{\rm spike}}^3 \tau_{Ca^{2+}}\mathrm{ln}(2)}{3}$, which makes the mean rate of NO production over time simply $\langle nNOS \rangle = \langle r \rangle \frac{{Ca^{2+}_{\rm spike}}^3 \tau_{Ca^{2+}} \mathrm{ln} (2)}{3}$. However, it is not sufficient to simply propose $nNOS_i(t) = r_i(t) \frac{{Ca^{2+}_{\rm spike}}^3 \tau_{Ca^{2+}} \mathrm{ln} (2)}{3}$. A single cell fires in the range of $\mathrm{3\,Hz}$. Therefore, the dynamics of $\rm nNOS$ are dominated by single spike events as shown in Figure \ref{nNOS_approx_plot} and it is---at least on the time scale of the observed oscillations---not a proper approximation to describe $\rm nNOS$ in this smooth, rate-based fashion.
\newpage
We introduced an idealized picture of diffusive homeostasis to deal with this, where diffusion across the tissue is instantaneous. This simplification results in a single level of NO-concentration for all neurons, only being modified over time by the sum of all neurons' NO-syntheses $nNOS_{\rm total}(t) \equiv \sum_{i} nNOS_i(t)$ and the decay term $-\lambda NO$. $nNOS_{\rm total}$ must be divided by the area $L^2$ of the tissue to account for the spread of NO. Furthermore, we introduce a random fluctuation term $\sigma_{NO} \xi(t)$ that accounts for local, momentary deviations from the mean. The idealized version of \eqref{NO_dyn} then reads
\begin{equation}
\dot{NO}(t)=-\lambda NO + \frac{nNOS_{\rm total}(t)}{L^2} + \sigma_{NO} \xi(t)\;.
\label{NO_dyn_approx}
\end{equation}
The amount of noise in the system is a hard-to-predict quantity since it depends on the spatial structure on the diffusive lattice, giving neurons that are further separated less impact onto each other. We tried to estimate the noise level based on the same instant-spread simplification made in Equation \eqref{NO_dyn_approx} to describe the overall NO-dynamics. This led to an underestimation of the total energy within the power spectrum in the final result. However, the noise level only acts as a global scaling factor of the shape of the power spectrum as long as we assume the noise in the system to be approximately white and was thus left open as a model fitting parameter.

It is clear from Equation \eqref{NO_dyn_approx} why we can now describe the \emph{total} influx of NO as the sum of a mean and a random fluctuation: While spiking events on individual cells remain sufficiently separated to justify equation \eqref{nNOS_single_sol_approx2}, the sum of all spike events from the excitatory population results in an effective Poisson process with a mean rate of $3 Hz \cdot N_{\rm exc.} = \mathrm{1200\,Hz}$, given the assumption that cells' spiking is approximately independent. We furthermore argue that a total rate of $\mathrm{1200\,Hz}$ is sufficiently large to express this rate as a continuous function of time, $r_{\rm total} = N_{\rm exc.} r_{\rm exc.pop.}(t)$, at least on the time scale of the oscillations to be studied (see  Figure \ref{rate_th_close}). We thus propose
\begin{equation}
nNOS_{\rm total}(t) = r_{\rm exc.pop.} N_{\rm exc.}\frac{{Ca^{2+}_{\rm spike}}^3 \tau_{Ca^{2+}} \mathrm{ln} (2)}{3}
\label{nNOS_total_dyn}
\end{equation} 
where $\frac{{Ca^{2+}_{\rm spike}}^3 \tau_{Ca^{2+}} \mathrm{ln} (2)}{3}$ is the integral from Equation \eqref{nNOS_single_sol_approx1}.

We wanted to simplify the dynamics of threshold adaption and its effect on the population rate as a next step. Corresponding to the reduction to a single variable $r_{\rm exc.pop.}(t)$ for the mean excitatory activity in equation \eqref{nNOS_total_dyn}, we wanted to find an appropriate description containing only the mean threshold $V_{t,\rm pop.}(t) = \langle V_t(t) \rangle$. As stated earlier, Figure \ref{rate_th_close} suggests an immediate and approximately linear relation between excitatory population rate and the mean of thresholds within the excitatory population. Figure \ref{thresh_r_linfit} shows a linear fit (being of the form $\alpha x + \beta$) of these two quantities plotted against each other. We found a good fit to the linear model apart from rarely appearing high values of $V_{t,\rm pop.}$. We thus expressed the population rate by $r_{\rm exc.pop.}(t) = r_{IP} + \alpha(V_{t,\rm pop.}(t)-V_{t,0})$, where $r_{IP}$ is given in Table \ref{Params_IP} as the mean excitatory target firing rate and $V_{t,0} \equiv (r_{IP}-\beta)/\alpha$ is the mean threshold corresponding to the target rate.
\begin{figure}
%%% linfit_th_rate.py
\includegraphics[width=\textwidth]{../../plots/Spike_Stats/linfit_th_rate.png}
\caption{Linear fit of excitatory population activity and mean excitatory threshold. Slope $\alpha$ of the fit: $\mathrm{-0.79\, Hz/mV}$. Offset $\beta$: $\mathrm{-41.97\, Hz}$. Mean squared error: $\mathrm{R^2 = 0.68}$.}
\label{thresh_r_linfit}
\end{figure}

@@

In the plastic network, the absence of this breaking of spatial homogeneity even led to an under-representation of bidirectional connections, which is known to be an effect of STDP in recurrent networks \cite{Syn_Plast_Abbott}. The separation between simulations with and without a spatial connection profile is retained after having implemented diffusive homeostasis, as shown in Figure \ref{RC_CFB_plot}. 
\begin{figure}
%%% plot_RC_CFB.py
\includegraphics[width=\textwidth]{../../plots/syn_topology/RC_CFB_plot.png}
\caption{Fraction of bidirectional connections normalized over the expected fraction for a random graph with equal total connection fraction. Data from 10 trials, linewidth represents standard error.}
\label{RC_CFB_plot}
\end{figure}
\subsubsection{Distribution of Synaptic Weights} \label{Syn_Weight_Dist_Section}
A well-studied property of neural networks is the distribution of synaptic weights. Experimental studies in hippocampus and cortical regions suggest heavy-tailed log-normal-like distributions of excitatory synaptic efficacies \cite{Song_Connectivity_2005,Lisman_Synapses_1993,Yasumatsu_Synapses_2008,Loewenstein_Spine_Sizes}. Theoretical models are mainly based on a combination of multiplicative and additive weight dynamics \cite{Loewenstein_Spine_Sizes,Statman_Synapses_2014}, which is in line with our implementation of multiplicative normalization and additive STDP. Figure \ref{Weight_Dist} compares the resulting distributions. We found that we can retain a log-normal-like distribution in combination with diffusive homeostasis. A slightly larger variance can be observed. A possible explanation refers back to our observations in Figure \ref{STDP_change}: Regarding STDP as a combination of a multiplicative and an additive random process, a broader distribution of weights can be explained by a smaller potentiating effect, which is the case for diffusive homeostasis in the beginning of the simulation. Still, it seems unlikely that a possible effect of these differences is still present at $\mathrm{t=1500\,s}$.
\begin{figure}
%%% plot_weight_dist.py
\includegraphics[width=\textwidth]{../../plots/syn_topology/syn_weight_dist.png}
\caption{Distribution of decadic logarithm of excitatory synaptic weights at $\mathrm{t=1500\,s}$ for diffusive and non-diffusive homeostasis (both with distance-dependent connectivity). Gaussian fits returned $\mathrm{\mu_{\rm diff.}=-0.17\, \mathrm{log}_{10}(mV)}$, $\mathrm{\mu_{\rm non-diff.}=-0.087\, \mathrm{log}_{10}(mV)}$, $\mathrm{\sigma_{\rm diff.}=0.64\, log_{10}(mV)}$ and $\mathrm{\sigma_{non-diff.}=0.42\, log_{10}(mV)}$.}
\label{Weight_Dist}
\end{figure}

@@

\caption{Pearson correlation coefficient of predicted and measured firing rates (see Figure \ref{Rand_Matrix_Sol_vs_Sim}) versus diffusion constant. Line width depicts standard error.}

>>>>

\textbf{Short-Term Plasticity:} A short-term plasticity (STP) mechanism acting on recurrent excitatory connections was implemented as presented in \cite{Markram_STP} as an additional stabilization of network activity. It modulates the effective synaptic weights by multiplying the value stored in the weight matrix $w_{ji}$ by two dynamic variables $x$ and $u$, $w^{\rm effective}_{ji} = w_{ji}\cdot x \cdot u$, each synapse owning a pair $(x,u)$. The dynamics of these variables are given by:

\begin{equation}
\dot{x} = \frac{1-x}{\tau_d},\; \dot{u} = \frac{U-u}{\tau_f}
\label{STP_dynamics1}
\end{equation}
Each presynaptic spike furthermore causes a change of $x$ and $u$ by
\begin{equation}
x \rightarrow x - x\cdot u,\; u \rightarrow u + U(1-u)
\label{STP_dynamics2}
\end{equation}
This is a phenomenological model of short-term plasticity that models two aspects of synaptic transmission, namely short-term depression (STD) and short-term facilitation (STF). STD is an effect that subsumes different mechanisms involved in the depletion of available neurotransmitters in the presynaptic terminal due to presynaptic action potentials. This fraction of depletion is represented by the variable $x$. Consequently, it is reduced upon a presynaptic spike. STF is based on the fact that a presynaptic action potential causes an increase of neurotransmitters available for release due to the influx of calcium into the presynaptic terminal \cite{Katz_1967,Katz_1968}. This effect is modeled by the variable $u$. The amount of potentially available neurotransmitters during an ``idle" state, with no presynaptic action potentials arriving, is given by $U$. Consequently, the dynamical system given by Equations \eqref{STP_dynamics1} and \eqref{STP_dynamics2} has its fixed point at $(x=1,u=U)$. If spikes arrive, depending on the choice of $\tau_d$ and $\tau_f$, one can achieve a weight modulation that is dominated by potentiation ($\tau_f \gg \tau_d$) or depression ($\tau_f \ll \tau_d$). We chose $U=0.04$, $\tau_d = 0.5s$ and $\tau_f = 2s$ as a rough approximation of the values that were experimentally observed \cite{Markram_STP}, giving it a tendency towards potentiation. However, one should keep in mind that for $U\in [0,1]$, $x\cdot u \in [0,1]$ always holds, thus the factor $x\cdot u$ has a generally diminishing effect. For our choice of variables for example, a Poisson input with a constant rate achieves the best synaptic transmission at a rate of $\sim 4.5 Hz$, corresponding to $x\cdot u \approx 0.2$. \textit{Potentiation} in this context refers to the fact that stronger input strengthens synaptic transmission \emph{compared} to close to zero incoming spikes.

@@

\newpage
\section{Results} \label{results}
We first present the results of a comparison between features of the network activity in both homeostatic variants in Section \ref{activ_analys}, since our main goal of the implementation of diffusive homeostasis was to allow the network to develop a broader distribution of firing rates across excitatory neurons compared to the original version of homeostasis. This is followed by an analytic discussion of an instability we observed within the diffusive homeostatic feedback loop, see Section \ref{theor_osc}. Furthermore, we compare topological features of the network under the influence of diffusive and non-diffusive homeostasis in section \ref{Network_Topology_Section}.

\subsection{Activity Analysis} \label{activ_analys}
As a first step, we investigated how diffusive homeostasis compares to non-diffusive homeostasis in terms of maintaining a stable level of population activity among excitatory and inhibitory population (though the latter one is only indirectly affected by excitatory synaptic input).
\subsubsection{Dynamic Behavior of the Population Activity}
We set the time constant of threshold adaptation in the diffusive homeostasis to $\mathrm{2500\, ms}$ on a first attempt, as given by \cite{Sweeney_Paper}. Switching between non-diffusive/diffusive homeostasis after $\rm 750\, s$, Figure \ref{full_sim_osci} shows the resulting dynamics of the population activity of the excitatory and inhibitory group (panel A/A*), the average $\rm NO$ concentration at the excitatory neurons' positions (panel B/B*) and the average firing threshold (panel C/C*) within the excitatory group (inhibitory threshold was fixed, see table \ref{LIF_neuron_params}). Both homeostatic mechanisms managed to keep the excitatory population activity in the desired range of 3 Hz. However, what might appear to be slightly faster and stronger random fluctuations in panel A-C were in fact regular oscillations across all three depicted variables, as can be seen in panel A*-C*, showing the dynamics in a smaller time window. While the oscillation amplitudes undergo a rather unpredictable course, the frequency remains at a constant level of $\simeq 0.5\, Hz$. This feature is also illustrated by the fact that the power spectra depicted in Figure \ref{Power_Spec_without_Analysis} have a peak at the frequency that appeared to be dominant in Figure \ref{full_sim_osci}. The fact that the mean over all power spectra of excitatory thresholds differs from the power spectrum of the mean of these thresholds in its amplitude indicates that not all thresholds oscillate at the same phase. The overall shapes of both spectra are still similar, though.
\begin{figure}
\begin{center}
%% oscillating_activity_plot.py
\includegraphics[width=\textwidth]{../../plots/diff_hom/rate_NO_th_compl.png}
\end{center}
%\includegraphics[width=\textwidth]{../../plots/diff_hom/rate_NO_th_closeup.png}
\caption{A/A*: Mean of excitatory (blue) and inhibitory (red) neuronal firing rate. B/B*: NO concentration averaged over readouts at excitatory neurons' positions. C/C*: Average excitatory firing threshold. Non-diffusive homeostasis was used for $\mathrm{0-700\, s}$ , diffusive h. for $\mathrm{700-1500\, s}$. Target activity was $\mathrm{3\, Hz}$.}
\label{full_sim_osci}
\end{figure}
\begin{figure}
\begin{center}
%%% plot_pow_spec.py
\includegraphics[width=0.7\textwidth]{../../plots/power_spec/power_spectrum_sim_only.png}
\caption{Blue: Mean of power spectrum over all excitatory thresholds. Red: Power spectrum of mean over all excitatory thresholds.}
\label{Power_Spec_without_Analysis}
\end{center}
\end{figure}

@@

We thus speculated whether the observed oscillations might be explicable based on the network topology and the dynamic interplay between excitatory and inhibitory population. For this purpose, we transferred the structure and dynamics onto a simpler, rate-based model of neural network activity. The results are discussed in the following section
\subsubsection{Predicting Activity in a Simplified Rate-Based Network}
Dynamical models of interconnected excitatory and inhibitory groups can exhibit sustained oscillations of activity on a larger, rate-based level \cite[p.~270]{Theor_Neur_Dayan}. We therefore considered this interplay as a source of the observed oscillations. To test this, we mapped the LIF-SORN onto a non-spiking neural network. We ignored plastic changes due to STDP, since we assumed these to be irrelevant in the context of the time scale of the observed oscillations.

As a simplest model, one can describe the dynamics of the excitatory and inhibitory populations as a two-dimensional dynamical system, representing the respective population rates $r_e$ and $r_i$

@@

$\mathrm{\langle w_{AB} \rangle,\, \langle N_{\mathrm{in},AB} \rangle}$ and $\mathrm{\langle r_A \rangle}$ denote the mean weight of a synapse connecting population $\mathrm{A}$ with population $\mathrm{B}$, the mean number of incoming connections at a neuron in population $\mathrm{B}$ and the mean firing rate in population $\mathrm{A}$ (which makes it necessary to assume changes within the firing rate to be slow enough for a quasi-stationary description of the activity).

@@

One thus needs to incorporate the factor $\mathrm{x_{st.}\cdot u_{st.}}$ by means of the previously described mean input and the contribution to the total variance of the membrane noise when trying to find the fixed point of Equations \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2}.

All these aspects taken together make it impossible to find an analytic expression for the fixed point of Equation \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2}. Running a simulation of Equations \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2}, including all aforementioned approximations resulted in a stable configuration. We set the relevant parameters to their predefined values for this simulation and additionally extracted the mean threshold of the excitatory population which was set by the non-diffusive homeostasis, see C/C* in Figure \ref{full_sim_osci}, yielding a threshold of $\mathrm{-56.963\, mV}$. The results are depicted in Figure \ref{Dyn_Rate_Approx}. The system settled to a fixed point of $\mathrm{(r_{0,e},r_{0,i}) = (2.992\, Hz, 7.144\, Hz)}$, which is quite close to $\mathrm{(3\, Hz, 6.768\, Hz)}$, being the mean excitatory and inhibitory firing rate found between $\mathrm{100-750\, s}$ of the full simulation (i.e., during non-diffusive homeostasis).
\begin{figure}
%% fixed_point_net.py
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/approx_rates_EI.png}
\end{center}
\caption{Dynamics of equations \ref{2d-exc_inh_rate_1} (blue line) and \ref{2d-exc_inh_rate_2} (red line). Dotted lines mark the mean frequencies that were present in the full spiking network. Excitatory: 3 Hz, Inhibitory: 6,768 Hz. \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2} converged to 2.992 Hz and 7.144 Hz respectively.}
\label{Dyn_Rate_Approx}
\end{figure}
In principle, one could now further quantify the stability of the found fixed point by calculating the Jacobian matrix at $\mathrm{(r_{0,e},r_{0,i})}$. For the sake of the initial question, though, namely the source of oscillation, it is sufficient at this point to restrict oneself to a preliminary result: The recurrent network as such (without homeostasis) is stable under the given choice of parameters, assuming validity of the described rate model. Furthermore, on an empirical basis, the occurrence of oscillations appears to be rather dependent on the choice of homeostasis since all parameters of the simulation were kept unchanged upon the transition after 750 s, apart from the homeostatic mechanism itself. Therefore, we present in the following section our further analysis of the dynamics underlying the diffusive homeostatic feedback.

@@

Figure \ref{nNOS_approx_plot} compares the approximation given by \eqref{nNOS_single_sol_approx2} to the full NO production model (Equations \eqref{Ca_dyn} and \eqref{nNOS_dyn}). Spikes were drawn from a Poisson process at a rate of 3 Hz. The simplified model fits very well for sufficiently isolated spike events, as predicted. One can observe a slightly worse but acceptable agreement for the rare event of two subsequent spikes appearing very close to each other, as seen in Figure \ref{nNOS_approx_plot} at approximately $\rm 1.5\, s$.

@@

\caption{Time course of nNOS(t) with Poisson spiking at 3 Hz. The full simulation (blue, see Equations \eqref{Ca_dyn},\eqref{nNOS_dyn}) is well fitted by the simplified model (red, see Equation \eqref{nNOS_single_sol_approx2}). Top panel is a closeup of the first spike event. Dashed lines mark the time span shown in the closeup.}

@@

Returning to the question of how to describe NO synthesis in a simplified form, we note that according to Equation \eqref{nNOS_single_sol_approx2}, a single spike causes the release of NO by an amount of $\frac{{Ca^{2+}_{\rm spike}}^3 \tau_{Ca^{2+}}\mathrm{ln}(2)}{3}$, which makes the mean rate of NO production over time simply $\langle nNOS \rangle = \langle r \rangle \frac{{Ca^{2+}_{\rm spike}}^3 \tau_{Ca^{2+}} \mathrm{ln} (2)}{3}$. However, it is not sufficient to simply propose $nNOS_i(t) = r_i(t) \frac{{Ca^{2+}_{\rm spike}}^3 \tau_{Ca^{2+}} \mathrm{ln} (2)}{3}$. A single cell fires in the range of $\mathrm{3\,Hz}$. Therefore, the dynamics of $\rm nNOS$ are dominated by single spike events as shown in Figure \ref{nNOS_approx_plot} and it is---at least on the time scale of the observed oscillations---not a proper approximation to describe $\rm nNOS$ in this smooth, rate-based fashion.
\newpage
We introduced an idealized picture of diffusive homeostasis to deal with this, where diffusion across the tissue is instantaneous. This simplification results in a single level of NO-concentration for all neurons, only being modified over time by the sum of all neurons' NO-syntheses $nNOS_{\rm total}(t) \equiv \sum_{i} nNOS_i(t)$ and the decay term $-\lambda NO$. $nNOS_{\rm total}$ must be divided by the area $L^2$ of the tissue to account for the spread of NO. Furthermore, we introduce a random fluctuation term $\sigma_{NO} \xi(t)$ that accounts for local, momentary deviations from the mean. The idealized version of \eqref{NO_dyn} then reads
\begin{equation}
\dot{NO}(t)=-\lambda NO + \frac{nNOS_{\rm total}(t)}{L^2} + \sigma_{NO} \xi(t)\;.
\label{NO_dyn_approx}
\end{equation}
The amount of noise in the system is a hard-to-predict quantity since it depends on the spatial structure on the diffusive lattice, giving neurons that are further separated less impact onto each other. We tried to estimate the noise level based on the same instant-spread simplification made in Equation \eqref{NO_dyn_approx} to describe the overall NO-dynamics. This led to an underestimation of the total energy within the power spectrum in the final result. However, the noise level only acts as a global scaling factor of the shape of the power spectrum as long as we assume the noise in the system to be approximately white and was thus left open as a model fitting parameter.

It is clear from Equation \eqref{NO_dyn_approx} why we can now describe the \emph{total} influx of NO as the sum of a mean and a random fluctuation: While spiking events on individual cells remain sufficiently separated to justify Equation \eqref{nNOS_single_sol_approx2}, the sum of all spike events from the excitatory population results in an effective Poisson process with a mean rate of $3 Hz \cdot N_{\rm exc.} = \mathrm{1200\,Hz}$, given the assumption that cells' spiking is approximately independent. We furthermore argue that a total rate of $\mathrm{1200\,Hz}$ is sufficiently large to express this rate as a continuous function of time, $r_{\rm total} = N_{\rm exc.} r_{\rm exc.pop.}(t)$, at least on the time scale of the oscillations to be studied (see  Figure \ref{rate_th_close}). We thus propose
\begin{equation}
nNOS_{\rm total}(t) = r_{\rm exc.pop.} N_{\rm exc.}\frac{{Ca^{2+}_{\rm spike}}^3 \tau_{Ca^{2+}} \mathrm{ln} (2)}{3}
\label{nNOS_total_dyn}
\end{equation} 
where $\frac{{Ca^{2+}_{\rm spike}}^3 \tau_{Ca^{2+}} \mathrm{ln} (2)}{3}$ is the integral from Equation \eqref{nNOS_single_sol_approx1}.

We wanted to simplify the dynamics of threshold adaption and its effect on the population rate as a next step. Corresponding to the reduction to a single variable $r_{\rm exc.pop.}(t)$ for the mean excitatory activity in Equation \eqref{nNOS_total_dyn}, we wanted to find an appropriate description containing only the mean threshold $V_{t,\rm pop.}(t) = \langle V_t(t) \rangle$. As stated earlier, Figure \ref{rate_th_close} suggests an immediate and approximately linear relation between excitatory population rate and the mean of thresholds within the excitatory population. Figure \ref{thresh_r_linfit} shows a linear fit (being of the form $\alpha x + \beta$) of these two quantities plotted against each other. We found a good fit to the linear model apart from rarely appearing high values of $V_{t,\rm pop.}$. We thus expressed the population rate by $r_{\rm exc.pop.}(t) = r_{IP} + \alpha(V_{t,\rm pop.}(t)-V_{t,0})$, where $r_{IP}$ is given in Table \ref{Params_IP} as the mean excitatory target firing rate and $V_{t,0} \equiv (r_{IP}-\beta)/\alpha$ is the mean threshold corresponding to the target rate.
\begin{figure}
%%% linfit_th_rate.py
\includegraphics[width=\textwidth]{../../plots/Spike_Stats/linfit_th_rate.png}
\caption{Linear fit of excitatory population activity and mean excitatory threshold. Slope $\alpha$ of the fit: $\mathrm{-0.79\, Hz/mV}$. Offset $\beta$: $\mathrm{-41.97\, Hz}$. Mean squared error: $\mathrm{R^2 = 0.68}$.}
\label{thresh_r_linfit}
\end{figure}

@@

In the plastic network, the absence of this breaking of spatial homogeneity even led to an under-representation of bidirectional connections, which is known to be an effect of STDP in recurrent networks \cite{Syn_Plast_Abbott}. The separation between simulations with and without a spatial connection profile is retained after having implemented diffusive homeostasis, as shown in Figure \ref{RC_CFB_plot}. 
\begin{figure}
%%% plot_RC_CFB.py
\includegraphics[width=\textwidth]{../../plots/syn_topology/RC_CFB_plot.png}
\caption{Fraction of bidirectional connections normalized over the expected fraction for a random graph with equal total connection fraction. Data from 10 trials, linewidth represents standard error.}
\label{RC_CFB_plot}
\end{figure}
\subsubsection{Distribution of Synaptic Weights} \label{Syn_Weight_Dist_Section}
A well-studied property of neural networks is the distribution of synaptic weights. Experimental studies in hippocampus and cortical regions suggest heavy-tailed log-normal-like distributions of excitatory synaptic efficacies \cite{Song_Connectivity_2005,Lisman_Synapses_1993,Yasumatsu_Synapses_2008,Loewenstein_Spine_Sizes}. Theoretical models are mainly based on a combination of multiplicative and additive weight dynamics \cite{Loewenstein_Spine_Sizes,Statman_Synapses_2014}, which is in line with our implementation of multiplicative normalization and additive STDP. Figure \ref{Weight_Dist} compares the resulting distributions. We found that we can retain a log-normal-like distribution in combination with diffusive homeostasis, giving a similar distribution as for non-diffusive homeostasis.
\begin{figure}
%%% plot_weight_dist_mult_sets.py
\includegraphics[width=\textwidth]{../../plots/syn_topology/syn_weight_dist_mult_sets.png}
\caption{Distribution of decadic logarithm of excitatory synaptic weights at $\mathrm{t=1500\,s}$ for diffusive and non-diffusive homeostasis (both with distance-dependent connectivity). Gaussian fits returned $\mathrm{\mu_{\rm diff.}=-0.128\, \mathrm{log}_{10}(mV)}$, $\mathrm{\mu_{\rm non-diff.}=-0.064\, \mathrm{log}_{10}(mV)}$, $\mathrm{\sigma_{\rm diff.}=0.415\, log_{10}(mV)}$ and $\mathrm{\sigma_{non-diff.}=0.395\, log_{10}(mV)}$. Weight matrices were taken from 5 simulations, each at $\rm t = 1500\,s$.}
\label{Weight_Dist}
\end{figure}

@@

\caption{Pearson correlation coefficient of predicted and measured firing rates (see Figure \ref{Rand_Matrix_Sol_vs_Sim}) versus diffusion constant. Line width depicts standard error, 5 simulations were used for each $\rm D$-value.}
